---
title: "532 Project"
author: "Seoyeon Park"
date: "3/28/2023"
output: html_document
---

```{r}
rm(list = ls())
```

Import libraries:
```{r}
library(survival)
library(ggplot2)
library(dplyr)
library(ggcorrplot)
library(KMsurv)
library(survminer)
library(GGally)
library(psych)
library(corrplot)
library(stats)
library(grid)
library(caret)
library(gtable)
```

```{r}
data(heart)
```

#Data

From https://rdrr.io/cran/survival/man/heart.html

jasa: original data

birth.dt:	birth date
accept.dt:	acceptance into program
tx.date:	transplant date
fu.date:	end of followup
fustat:	dead or alive
surgery:	prior bypass surgery
age:	age (in years)
futime:	followup time
wait.time:	time before transplant
transplant:	transplant indicator (0 = no, 1 = yes)
mismatch:	mismatch score
hla.a2:	particular type of mismatch
mscore:	another mismatch score
reject:	rejection occurred


Use the already processed data, heart for EDA:

start, stop, event:	Entry and exit time and status for this interval of time
age:	age - 48 years
year:	year of acceptance (in years after 1 Nov 1967)
surgery:	prior bypass surgery 1=yes
transplant:	received transplant 1=yes
id:	patient id

Patients that identified to be eligible for a heart transplant were followed until death or censorship 
Original goal of the study was to assess whether patients receiving transplants survived longer than patients who did not receive a transplant

## heart

```{r}
heart
```

```{r}
summary(heart)
```


```{r}
as.factor(heart$surgery)
```


```{r}
summary(heart)
```

```{r}
min(heart$age + 48) #as young as 8 years old
max(heart$age + 48) #max age is 64
```

```{r}
unique(heart$id)
```

```{r}
table(jasa$transplant)
```

```{r}
table(jasa$surgery)
```



## jasa

```{r}
summary(jasa)
```


```{r}
summary(as.factor(jasa$transplant))
```

```{r}
summary(as.factor(jasa$surgery))
```

```{r}
summary(is.na(jasa))
summary(jasa)
```


## EDA

```{r}
ggplot(heart, aes(x=age + 48))+ geom_histogram(binwidth=5, colour="white", fill="#FF6666", alpha=0.8) + scale_x_continuous(breaks=seq(40,100,10)) +labs(title="Age Distribution") + theme_minimal() + xlab("Age") + ylab("Counts") + ggtitle("Age Distribution") 
ggsave("age distribution.jpg")
```


Look at 6 different age groups:
[5,15) [15,25) [25,35) [35,45) [45,55) [55,65)

```{r}
cut(heart$age + 48, breaks = c(5, 15, 25, 35, 45, 55, 65), right = F)
```

```{r}
summary(cut(heart$age + 48, breaks = c(5, 15, 25, 35, 45, 55, 65), right = F))
```

```{r}
pie_data <- data.frame(group = c("[5,15)", "[15,25)", "[25,35)", "[35,45)", "[45,55)", "[55,65)"), value = c(1/172, 6/172, 15/172, 42/172, 94/172, 14/172))
pie_data$group <- factor(pie_data$group, levels = pie_data$group, ordered = T)
ggplot(pie_data, aes(x="", y=value, fill=group)) + geom_bar(stat="identity", width=1) + coord_polar("y", start=0)  + theme_minimal() + xlab("") + ylab("") + ggtitle("Proportion of Age Groups") 
ggsave("age pie.jpg")
```


```{r}
heart$surgery <- as.factor(heart$surgery)
heart$transplant <- as.factor(heart$transplant)
```

```{r}
ggplot(jasa, show.legend = FALSE) + geom_bar(mapping = aes(y = surgery, fill = as.factor(surgery))) + theme_minimal() + xlab("Counts") + ylab("Prior Bypass Surgery") + ggtitle("Proportion of Surgery") + labs(fill = "Prior Bypass Surgery, Yes = 1")
ggsave("surgery proportion.jpg")
```

```{r}
ggplot(jasa, show.legend = FALSE) + geom_bar(mapping = aes(y = transplant, fill = as.factor(transplant))) + theme_minimal() + xlab("Counts") + ylab("Transplant Received") + ggtitle("Proportion of Transplants") + labs(fill = "Transplant Received, Yes = 1")
ggsave("transplant proportion.jpg")
```

Maybe include this in analysis for deciding covariates (and looking at time-dependent covariates too)

```{r}
model.matrix(~0+., data=jasa) %>% 
  cor(use="pairwise.complete.obs") %>% 
  ggcorrplot(show.diag=FALSE, type="lower", lab=TRUE, lab_size=8)
```

"The key rule for time dependent covariates in a Cox model is simple and essentially the same as that for gambling: you cannot look into the future. A covariate may change in any way based on past data or outcomes, but it may not reach forward in time."

Anomalies in the data:
https://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf pg 10

In original data, a patient has died on the same day as acceptance into the program. 

```{r}
subset(jasa, accept.dt == fu.date)
```

in the heart dataset processed from the raw data jasa, it is important to note that some patients have multiple rows 

For example, patient id 3, 4, 7, etc. 

```{r}
heart
```

```{r}
data(heart)
```

## KME - Very basic approach!

Examine survival as a function of whether subject received a heart transplant, or has a history of prior bypass surgery

Difference in survival for transplant - use jasa
p.val = T gives a log-rank test (is there a difference in survival between groups)

```{r}
survTransplant <- survdiff(Surv(futime, fustat) ~ transplant, data = jasa, rho = 0)
survTransplant
```

```{r}
summary(survTransplant)
```


```{r}
survplotTransplant <- ggsurvplot(survfit(Surv(futime, fustat) ~ transplant, data = jasa), conf.int=TRUE, pval=TRUE, pval.method = T, surv.median.line = "hv",
           legend.labs=c("Transplant = No", "Transplant = Yes"), legend.title="Transplant",  
           palette=c("dodgerblue2", "orchid2"), 
           title="Kaplan-Meier Curve for Heart Transplant Survival", 
           risk.table.height=.15)
survplotTransplant
ggsave("survplotTransplant.jpg")
```

```{r}
survSurgery <- survfit(Surv(futime, fustat) ~ surgery, data = jasa, conf.type = "plain")
survSurgery
```

```{r}
survplotSurgery <- ggsurvplot(survSurgery, conf.int=TRUE, pval=TRUE, pval.method = T, surv.median.line = "hv",
           legend.labs=c("Prior Bypass Surgery = No", "Prior Bypass Surgery = Yes"), legend.title = "Surgery",  
           palette=c("dodgerblue2", "orchid2"), 
           title="Kaplan-Meier Curve for Heart Transplant Survival", 
           risk.table.height=.15)
survplotSurgery
ggsave("survplotSurgery.jpg")
```

Issues with doing a simple KME:
- Patients who received transplants had to wait from acceptance date on the transplant list until a suitable donor was found 
- During this wait period they were all at risk for death, but did not have a transplant
- Thus the wait time accumulated by the transplant patients contributes to survival of the non-transplant patients (Unless total survival time for each patient was used)

- tissue mismatch score and age transplant only are indicators for survival of the patients who did indeed receive a transplant 

## Cox-PH

exp(coef) gives the hazard ratio:
HR = 1, no effect
HR > 1, increase in hazard
HR < 1: Reduction in hazard


Because of the second issue of this problem we need to separate individual's survival before and after transplant.

Lucky for us, we already have a cleaned dataset that separated this already for us:
i.e. use heart NOT jasa

In the heart dataset, patients who received a transplant are treated as two cases, before and after the operation, so cases in the transplant group are in general both right censored and left-truncated. This is handled by Surv by supplying entry and exit times. 
For example, patient 3, 4, 7 

```{r}
jasa
```

```{r}
heart$surgery <- as.factor(heart$surgery)
heart$transplant <- as.factor(heart$transplant)
```


```{r}
heart
```

NOT the correct coxph model! 
```{r}
fit1 <- coxph(Surv(futime, fustat) ~ transplant, data = jasa)
fit1
```

Need to correct for time-dependent transplant status - i.e. use heart dataset

Fit all variables:
```{r}
heart
```


Full Cox PH:

```{r}
cox1 <- coxph(Surv(start, stop, event) ∼ age + year + surgery + transplant, data = heart)
cox1
```

```{r}
summary(cox1)
```

```{r}
temp <- cox.zph(cox1, transform = "log")
temp
```

```{r}
plot(temp[1])
abline(h = cox1$coefficients[1], lty = 4, col = "red")
```

```{r}
plot.new()
plot(temp[2])
abline(h = cox1$coefficients[2], lty = 4, col = "red")
```

```{r}
plot.new()
plot(temp[3])
abline(h = cox1$coefficients[3], lty = 4, col = "red")
```

```{r}
plot.new()
plot(temp[4])
abline(h = cox1$coefficients[4], lty = 4, col = "red")
```

A better model to consider would be to include an interaction of year and transplant. As well, stratify the transplant variable as it is time dependent. 

-> Transplant lists are in order, therefore the year of acceptance into the program definitely has an interaction with transplant. 

```{r}
cox2 <- coxph(Surv(start, stop, event) ∼ strata(transplant) + age + year + surgery + year:transplant, data = heart)
cox2
```

```{r}
summary(cox2)
```


```{r}
temp <- cox.zph(cox2, transform = "log")
temp
```


```{r}
plot(temp, conf.int = T)
```

```{r}
plot.new()
plot(temp[1])
abline(h = cox2$coefficients[1], lty = 4, col = "red")
```

```{r}
plot.new()
plot(temp[2])
abline(h = cox2$coefficients[2], lty = 4, col = "red")
```

```{r}
plot.new()
plot(temp[3])
abline(h = cox2$coefficients[3], lty = 4, col = "red")
```

```{r}
plot.new()
plot(temp[4])
abline(h = cox2$coefficients[4], lty = 4, col = "red")
```

For both models, n = 172 only refers to the number of observations in the dataset. There are duplicates of patient id in the heart data!












